---
title: Literature analysis
author: Francis Leblanc
date: '2020-08-01'
slug: literature-analysis
categories:
  - R
tags:
  - Atrial fibrillation
  - R
  - literature
subtitle: ''
summary: ''
authors: []
lastmod: '2020-08-12T15:12:01-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
toc: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	fig.height=7,
	fig.width=10,
	dpi = 200
)
```


Here I discuss the various literature packages in R and show a use case on a topic I study.
Available packages; 

* bibliometrix: has a variety of functions returning most useful metrics you could want
* scholar: useful to evaluate journals and scholars (here I use it to retrieve impact factors of journals)
* pubmed.mineR: allows to query PubMed in multiple ways (great option if your query results is very high, but it’s slow and you won’t get citation counts and pubtator data)
* wosr: If you have an API username and password, this might be the best option to query a large amount of papers
* other packages: pubmedR, RISmed and RISmed

After spending some time exploring the different options, I found that MeSH terms were often missing, and Keywords or Keywords plus were not very informative. I ended up finding the results from [pubtator](https://www.ncbi.nlm.nih.gov/research/pubtator/). This is a NCBI Deep learning algorithm that identified and categorised 5 class of words in publications (Genes, Diseases, Mutations, Chemicals, Species) from the PMID. 
For use cases where you have a large number of results, I provide a script at the end of the document. 




Aims: 

* Determin the Keywords associated to the 14q32 locus evolution in time
* Determin journals yielding most citations on the subject



The following data was retreived from Web of science database with the following search therms: **TOPIC: ((14q32)  OR (DLK1-DIO3)) AND TOPIC: (heart  OR cardi)** The results were exported as plain text files.


```{r}
pacman::p_load(bibliometrix,
               pubmed.mineR,
               rio,
               ttutils,
               tidyverse,
               ggplot2,
               ggrepel,
               scholar,
               reshape2)
```


# Overview of the results


```{r }
f <- c("data_litterature_14q32/14q32_OR_DLK1-DIO3_AND_heartORcardio.txt")
M <- convert2df(file = f, dbsource = "wos", format = "plaintext")
results <- biblioAnalysis(M, sep = ";")
plot(x = results, k = 10, pause = FALSE)
```

# Co-occurence of Author Keywords

```{r}
# Create keyword co-occurrences network
NetMatrix <- biblioNetwork(M, analysis = "co-occurrences", network = "author_keywords", sep = ";")

# Plot the network
net=networkPlot(NetMatrix, normalize="association", weighted=T, n = 30, Title = "Keyword Co-occurrences", type = "auto", size=T,edgesize = 5,labelsize=0.7)
```

# Genes and diseases analysis

Keywords are quite uninforamtive. I find that a better option is to use gene names and diseases identified by the NCBI Deep learning algorythm [pubtator](https://www.ncbi.nlm.nih.gov/research/pubtator/).

## Pubtator query

Pubator returns all identifiers merged from a PMID list. Here we want this information for each articles so we need to do this iteratively. 
```{r}
# pubtator_o <- list()
# for (i in 1:length(M$PM)) {
#   pubtator_o[[i]] <- pubtator_function(M$PM[i])
#   if (isInteger(i/100)) {
#     print(i)
#   }
# }
# export(pubtator_o, "data_litterature_14q32/14q32_OR_DLK1-DIO3_AND_heartORcardio_pubtator.rds")

pubtator_o <- import("data_litterature_14q32/14q32_OR_DLK1-DIO3_AND_heartORcardio_pubtator.rds")
```
## Network analysis of genes

Text needs to be reformated to be processed with the bibliometrix package

```{r}
do <- as.data.frame(do.call(rbind, lapply(pubtator_o, as.character)), stringsAsFactors = F)
colnames(do) <- names(pubtator_o[[1]])
do <- do %>% replace(.=="NULL", NA)
do <- do[do$PMID != " No Data ",]

do <- apply(do,2,function(x) gsub("c[(]", "", x))
do <- apply(do,2,function(x) gsub("[)]", "", x))
do <- as.data.frame(apply(do, 2,function(x) gsub("\"", "", x)), stringsAsFactors = F)
do <- as.data.frame(apply(do, 2,function(x) gsub(">[0-9]+","", x)), stringsAsFactors = F)
do <- as.data.frame(apply(do, 2,function(x) gsub(">MESH:[A-Z][0-9]+","", x)), stringsAsFactors = F)
do <- as.data.frame(apply(do, 2,function(x) gsub(">-","", x)), stringsAsFactors = F)
do <- as.data.frame(apply(do, 2,function(x) gsub(",",";", x)), stringsAsFactors = F)
```

<br>

Now we can replace Keywords with the columns for Genes and/or Diseases from Pubtator

```{r}
M2 <- M # new df to replace ID
df <- merge(M, do, by.x="PM", by.y = "PMID", all.x = T)


change_ID <- function(new_ID, DF) {
  # remove duplicated gene names (some are capitalized and others not)
  DF$ID <- new_ID
  DF$ID <- gsub("NA","",DF$ID)
  t <- split(toupper(DF$ID), seq(length(DF$ID)))
  t <- lapply(t, function(x) unique(str_trim(strsplit(x, ";")[[1]])))
  t <- lapply(t, function(x) paste(x, collapse = ";"))
  DF$ID <- unlist(t)
  DF$ID <- gsub("NULL","",DF$ID)
  return(DF)
}

new_ID <- paste0(df$Genes) # replacing ID with genes from Pubtator
M2 <- change_ID(new_ID,M2)

##### These other functions may also be helpful.
# summary(results2)
# CS <- conceptualStructure(M2,field="ID", method="CA", minDegree=4, clust=5, stemming=FALSE, labelsize=10, documents=10)

# rio::export(M2, "data_litterature_14q32/14q32_lit.xlsx") # to import in shiny
# biblioshiny()
```


```{r}
NetMatrix <- biblioNetwork(M2, analysis = "co-occurrences", network = "keywords", sep = ";")

net=networkPlot(NetMatrix, normalize="association", weighted=T, n = 30, Title = "Keyword Cooccurrences", type = "auto", size=T,edgesize = 5)
```


## Variation of investigated genes over time

```{r }
ID_evol_p <- function(DF, top_n_ID) {
  KWG_df <- KeywordGrowth(DF, Tag = "ID", sep = ";", top = top_n_ID, cdf = TRUE)
  m <- melt(KWG_df,  measure.vars = c(colnames(KWG_df)[2:(top_n_ID+1)]))
  m <- m %>% mutate(label = if_else(Year == max(Year), as.character(variable), NA_character_))
  
  labelInfo <-
    split(m, m$variable) %>%
    lapply(function(t) {
      data.frame(
        predAtMax = loess(value ~ Year, data = t) %>%
          predict(newdata = data.frame(Year = max(t$Year)))
        , max = max(t$Year)
      )}) %>%
    bind_rows
  
  labelInfo$label = levels(factor(m$variable))
  
  p <- ggplot(m, aes(x=Year, y=value, colour=variable))+
    geom_smooth(se=F) +
    geom_label_repel(data = labelInfo, 
                     aes(x = max, y = predAtMax, 
                         label = label, 
                         color = label),
                     size = 3,
                     nudge_x = 5) +
    theme_minimal() + 
    theme(legend.position = "none")+
    xlim(min(m$Year), 10+max(m$Year))+
    ylab("Cumulative frequency")
  return(p)
}

ID_evol_p(M2,30)
```

## Network analysis of diseases

```{r }
new_ID <- paste0(df$Diseases) # replacing ID with genes from Pubtator
M2 <- change_ID(new_ID,M2)
```


```{r}
NetMatrix <- biblioNetwork(M2, analysis = "co-occurrences", network = "keywords", sep = ";")

net=networkPlot(NetMatrix, normalize="association", weighted=T, n = 50, Title = "Keyword Cooccurrences", type = "auto", size=T,edgesize = 5,labelsize=0.7)
```

## Variation of investigated diseases over time

```{r}
ID_evol_p(M2,30)
```

## Combined genes and diseases network 

```{r}
new_ID <- paste0(df$Genes, ";",df$Diseases)
M2 <- change_ID(new_ID,M2)
results2 <- biblioAnalysis(M2, sep = ";")
# summary(results2)
# rio::export(M2, "data_litterature_14q32/14q32_lit.xlsx")
```


```{r}
NetMatrix <- biblioNetwork(M2, analysis = "co-occurrences", network = "keywords", sep = ";")

net=networkPlot(NetMatrix, normalize="association", weighted=T, n = 50, Title = "Keyword Cooccurrences", type = "auto", size=T,edgesize = 5,labelsize=0.7)
```

```{r}
library("wordcloud")
set.seed(1234)
d <- as.data.frame(results2$ID)
wordcloud(words = d$Tab, freq = d$Freq, min.freq = 2,
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))
```

# Best journal in topic

* number of citation per paper per year
* compare to their impact factor

```{r}
sources <- data.frame(sources = M2$SO , TCperYear = results2$TCperYear)
sources <- sources %>% group_by(sources) %>% summarise(TCpY = sum(TCperYear))
sources$n_aticles <- M2 %>% group_by(SO) %>% count() %>% pull(n)
sources$TCpYpA <- sources$TCpY/sources$n_aticles


impact <- get_impactfactor(journals=sources$sources, max.distance = 0.1)
sources <- merge(sources, impact, by.x="sources", by.y = "Journal", all.x = T)
sources <- sources[!duplicated(sources$sources),]
top_J <- head(sources[order(sources$TCpYpA, decreasing = T),], 15)
```

```{r}
ggplot(top_J, aes(y=reorder(sources,TCpYpA),x=TCpYpA))+
  geom_bar(stat = "identity") +
  theme_minimal()+
  theme(legend.position = "none")+
  xlab("Citations/(Years*Articles)")+
  ylab("Journal")
```

```{r fig.height=10, fig.width=10}
ggplot(top_J, aes(x = ImpactFactor, y=TCpYpA, label=sources, colour=sources))+
  geom_point() +
  geom_label_repel(size = 2.3)+
  scale_y_log10() +
  scale_x_log10() +
  theme_minimal() +
  theme(legend.position = "none") +
  ylab("Citations/(Years*Articles)")
```






# Large queries analyses

```{r eval=FALSE}
##### Function to get citation count from PMID 
##### (modified from RISmed package to avoid errors from multiple queries)
f <- function(id){
  
  base <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/elink.fcgi?retmode=xml&dbfrom=pubmed&id=ID&cmd=neighbor"
  base <- sub("ID", id, base)
  
  the_url <- url(description = base)
  on.exit(close(the_url))
  
  lines <- readLines(the_url)
  
  citedstart <- grep("pubmed_pubmed_citedin", lines)
  
  if(length(citedstart) == 0){
    return(0)
  }
  else{
    citedend <- grep("pubmed_pubmed", lines)
    citedend <- min(citedend[citedend > citedstart])
    
    tags <- grep("<Id>", lines)
    tags <- tags[tags > citedstart & tags < citedend]
    
    if(any(tags)){
      hits <- sub("(.*<ID>)([0-9]+)(.*)","\\2",lines[tags])
      hits <- unique(hits)
      length(hits[hits != id])
    }
    else{
      0
    }	
  }
}

# get 
library(pubmedR)
query <- "atrial fibrillation[Title/Abstract]"
res <- pmQueryTotalCount(query = query, api_key = NULL)
D <- pmApiRequest(query = query, limit = 100000, api_key = NULL)
M <- pmApi2df(D)


library(ttutils)
v <- c()
for (i in 1:length(M$PMID)) {
  v <- c(v,f(M$PMID[i])) # get citations 1 by one with sleep to avoid errors
  Sys.sleep(0.3)
  if (isInteger(i/100)) { # follow the long process
    print(i)
  }
}

# get Genes, Diseases, Mutations, Chemicals, Species from pubtator
library(pubmed.mineR)
pubtator_o <- list()
for (i in 1:length(M$PMID[1:100])) {
  pubtator_o[[i]] <- pubtator_function(M$PMID[i])
  if (isInteger(i/100)) {
    print(i)
  }
}
```




```{r}
sessionInfo()
```

